Advancing spatially explicit fisheries management with age-structured species distribution models

### 2025-01-29 paper.qmd
tinkering with YAML

---
title: "Advancing spatially explicit fisheries management with age-structured species distribution models"
format: pdf
crossref:
  custom:
    - kind: float
      key: supfig
      latex-env: supfig
      reference-prefix: Fig. S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
author:
  - name: Eric Ward
    affiliation:
      - id: cb
        name: Consercation Biology Division, Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanographic and Atmospheric Administration
        address: 2725 Montlake Blvd E
        city: Seattle
        state: WA
        zip: 98112
  - name: Nick Tolimieri
    affiliation:
      - ref: cb

@@

Estimates of age composition by year may be useful in tracking the distribution of cohorts through time, as well as examining variability in distributions of particular age classes over time. For example, for both hake ([Figs. @fig-hake-spatial-composition] & [@supfig-hake-spatial-composition-all]) and sablefish ([Figs. @fig-sablefish-spatial-composition] & [@supfig-sablefish-spatial-composition-all]), distinct patches of high age-0 abundance tended to disperse as the fish aged such that spatial distributions for older age-classes were less distinct. Additionally, results from our models can be used to project the spatial distributions of strong hake cohorts (2008, 2010, 2014, 2016; [Fig. @fig-hake-spatial-composition]) and strong sablefish cohorts (2008, 2010, 2016, 2021, [Fig. @fig-sablefish-spatial-composition]). These predictions are also useful in revealing subtle differences between strong cohort years. For example, the 2014 cohort of Pacific hake had a larger fraction of age-1 individuals distributed in the northern part of the domain (along the coast of Washington state), while other cohorts had hotspots in the center and southern portion of the domain. Similarly for sablefish, the 2016 cohort was concentrated in the north (age-1 individuals in 2011), versus other cohorts that had a more coastwide distribution. In fact, age-1 fish were uncommon south of approximately Cape Mendocino (40 °N) from 2015 - 2020 ([@supfig-sablefish-spatial-composition-all]), concurrent with 'the Blob' and other marine heatwaves.

@@

Our 1-step ahead validation analyses indicate that there is a stronger association between older ages compared to younger ages, and this effect is strongest for sablefish ([Fig. @fig-glm-coefficients]). When comparing our initial Poisson GLM to a more complicated GLMM with random year effects in the intercept and slope, we found little support for shared correlations across age classes of a given species. Shared correlations would indicate for example that age-2 and age-3 sablefish may be more or less predictable in certain years (indicating that information about one age class may be useful in predicting another). With the exception of a strong positive correlation between age-0 and age-1 sablefish, most correlations between age classes were smaller than 0.3 in magnitude ([@supfig-glm-coefficients-time]).

@@

Our age-structured SDM framework represents a significant advance in spatial modeling, offering novel insights into the dynamics of fish populations and their age-specific distributions. From a management perspective, age-structured SDMs offer a powerful tool for addressing critical challenges, including bycatch mitigation, spatial planning, and sustainable harvesting. By providing high-resolution predictions of spatial age composition, these models enable managers to align fishing effort with ecological and economic goals, ensuring the long-term sustainability of marine resources. Future research could extend this framework by incorporating environmental covariates, such as temperature or oxygen, to account for dynamic habitat preferences, or fisheries removals to account for local depletion [@ono2016]. Additionally, exploring the potential for multi-species models could provide insights into ecosystem-level interactions and their influence on spatial distributions.

@@

```{=tex}
\break
\clearpage
```
## Supplementary material {.appendix}

```{r label="supfig-spatial-parameters", fig.cap="Estimated spatial parameters (range, and spatial and spatiotemporal standard deviations). Points represent mean estimates; lines represent 95% confidence intervals", echo=FALSE}
knitr::include_graphics("plots/spatial_parameters.png")
```

@@

```{r label="supfig-hake-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent ages (1-5). CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}
knitr::include_graphics("plots/Pacific hake-age-class-year.png")
```

@@

```{r label="supfig-sablefish-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent ages (0-4). CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}

@@

```{r label="supfig-glm-coefficients-time", fig.cap="Time-varying coefficients (with random effects in year) relating predicted densities of age a fish in year t to observed numbers the following year. ", echo=FALSE}
knitr::include_graphics("plots/glm_coefficients_time.png")
```

>>>>

---
title: "Advancing spatially explicit fisheries management with age-structured species distribution models"

format: 
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
author:
  - name: Eric Ward
  - name: Nick Tolimieri
    affiliations:
    - Conservation Biology Division, Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanographic and Atmospheric Administration, 2725 Montlake Blvd E, Seattle WA 98112

@@

Estimates of age composition by year may be useful in tracking the distribution of cohorts through time, as well as examining variability in distributions of particular age classes over time. For example, for both hake ([Figs. @fig-hake-spatial-composition] & [-@fig-hake-spatial-composition-all]) and sablefish ([Figs. @fig-sablefish-spatial-composition] & [-@fig-sablefish-spatial-composition-all]), distinct patches of high age-0 abundance tended to disperse as the fish aged such that spatial distributions for older age-classes were less distinct. Additionally, results from our models can be used to project the spatial distributions of strong hake cohorts (2008, 2010, 2014, 2016; [Fig. @fig-hake-spatial-composition]) and strong sablefish cohorts (2008, 2010, 2016, 2021, [Fig. @fig-sablefish-spatial-composition]). These predictions are also useful in revealing subtle differences between strong cohort years. For example, the 2014 cohort of Pacific hake had a larger fraction of age-1 individuals distributed in the northern part of the domain (along the coast of Washington state), while other cohorts had hotspots in the center and southern portion of the domain. Similarly for sablefish, the 2016 cohort was concentrated in the north (age-1 individuals in 2011), versus other cohorts that had a more coastwide distribution. In fact, age-1 fish were uncommon south of approximately Cape Mendocino (40 °N) from 2015 - 2020 ([@fig-sablefish-spatial-composition-all]), concurrent with 'the Blob' and other marine heatwaves.

@@

Our 1-step ahead validation analyses indicate that there is a stronger association between older ages compared to younger ages, and this effect is strongest for sablefish ([Fig. @fig-glm-coefficients]). When comparing our initial Poisson GLM to a more complicated GLMM with random year effects in the intercept and slope, we found little support for shared correlations across age classes of a given species. Shared correlations would indicate for example that age-2 and age-3 sablefish may be more or less predictable in certain years (indicating that information about one age class may be useful in predicting another). With the exception of a strong positive correlation between age-0 and age-1 sablefish, most correlations between age classes were smaller than 0.3 in magnitude ([@fig-glm-coefficients-time]).

@@

Our age-structured SDM framework represents a significant advance in spatial modeling, offering novel insights into the dynamics of fish populations and their age-specific distributions. From a management perspective, age-structured SDMs offer a powerful tool for addressing critical challenges, including bycatch mitigation, spatial planning, and sustainable harvesting. By providing high-resolution predictions of spatial age composition, these models enable managers to align fishing effort with ecological and economic goals, ensuring the long-term sustainability of marine resources. Future research could extend this framework by incorporating environmental covariates, such as temperature or oxygen, to account for dynamic habitat preferences, or fisheries removals to account for local depletion [@ono2016space]. Additionally, exploring the potential for multi-species models could provide insights into ecosystem-level interactions and their influence on spatial distributions.

@@

```{=tex}
\break
\clearpage
```
# Supplementary material {.appendix}

```{r }
#| label: fig-spatial-parameters
#| echo: false
#| fig-cap: "Estimated spatial parameters (range, and spatial and spatiotemporal standard deviations). Points represent mean estimates; lines represent 95% confidence intervals"

knitr::include_graphics("plots/spatial_parameters.png")
```

@@

```{r label="fig-hake-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent ages (1-5). CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}
knitr::include_graphics("plots/Pacific hake-age-class-year.png")
```

@@

```{r label="fig-sablefish-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent ages (0-4). CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}

@@

```{r label="fig-glm-coefficients-time", fig.cap="Time-varying coefficients (with random effects in year) relating predicted densities of age a fish in year t to observed numbers the following year. ", echo=FALSE}
knitr::include_graphics("plots/glm_coefficients_time.png")
```
